---
title: "Comparison between `ALEPlot` and `ale` packages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison between `ALEPlot` and `ale` packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r general setup}
library(dplyr)
```

The [`ALEPlot` package](https://CRAN.r-project.org/package=ALEPlot) is the reference implementation of the brilliant idea of [accumulated local effects (ALE) by Daniel Apley and Jingyu Zhu](https://www.doi.org/10.1111/rssb.12377). However, it has not been updated since 2018. The [`ale` package](ale-intro.html "Introduction to the ale package") attempts to rewrite and extend the original base work.

In developing the `ale` package, we must ensure that we correctly implement the original ALE algorithm while extending it. Indeed, some permanent unit tests call `ALEPlot` to make sure that `ale` provides identical results for identical inputs. We thought that presenting some of these comparisons as a vignette might be helpful. We focus here on the [examples from the `ALEPlot` package](https://CRAN.r-project.org/package=ALEPlot/vignettes/AccumulatedLocalEffectPlot.pdf) so that results are directly comparable.

Other than its [extensions for ALE-based statistics](ale-statistics.html), here are some of the main points in which `ale` differs from `ALEPlot` where it provides otherwise similar functionality:

-   It uses `ggplot2` instead of base R graphics. We consider `ggplot2` to be a more modern and versatile graphics system.
-   It saves plots as ggplot objects to a "plots" element of the return value; it does not automatically print the plot to the screen. As we show, this lets the user manipulate the plots more flexibly.
-   In the plot, the Y outcome variable is displayed by default on its full absolute scale, centred on the median or mean, not on a scale relative to zero. (This option can be controlled with the `relative_y` argument, as we demonstrate.) We believe that such plots more easily interpretable.
-   In addition, there are numerous design choices to simply the function interface based on [tidyverse](https://www.tidyverse.org/) design principles.

One notable difference between the two packages is that the `ale` package does not and will not implement partial dependency plots (PDP). The package is focused exclusively on accumulated local effects (ALE); users who need PDPs may use `ALEPlot` or other implementations.

In each section here, we cover an example from `ALEPlot` and then reimplement it with `ale`.

## Simulated data with numeric outcomes (ALEPlot Example 2)

We begin with the second code example directly from the `ALEPlot` package. (We skip the first example because it is a subset of the second, simply without interactions.) Here is the code from the example to create a simulated dataset and train a neural network on it:

```{r nnet setup, fig.width=10, fig.height=8}
## R code for Example 2
## Load relevant packages
library(ALEPlot)
library(nnet)

## Generate some data and fit a neural network supervised learning model
set.seed(0)  # not in the original, but added for reproducibility
n = 5000
x1 <- runif(n, min = 0, max = 1)
x2 <- runif(n, min = 0, max = 1)
x3 <- runif(n, min = 0, max = 1)
x4 <- runif(n, min = 0, max = 1)
y = 4*x1 + 3.87*x2^2 + 2.97*exp(-5+10*x3)/(1+exp(-5+10*x3))+
13.86*(x1-0.5)*(x2-0.5)+ rnorm(n, 0, 1)
DAT <- data.frame(y, x1, x2, x3, x4)
nnet.DAT <- nnet(y~., data = DAT, linout = T, skip = F, size = 6,
decay = 0.1, maxit = 1000, trace = F)
```

For the demonstration, `x1` has a linear relationship with `y`, `x2` and `x3` have non-linear relationships, and `x4` is a random variable with no relationship with `y`. `x1` and `x2` interact with each other in their relationship with `y`.

### ALEPlot code

To create ALE data and plots, `ALEPlot` requires the creation of a custom prediction function:

```{r ALEPlot nnet yhat}
## Define the predictive function
yhat <- function(X.model, newdata) as.numeric(predict(X.model, newdata,
type = "raw"))

```

Now the `ALEPlot` function can be called to create the ALE data and plot it. The function returns a specially formatted list with the ALE data; it can be saved for subsequent custom plotting.

```{r ALEPlot nnet one-way}
## Calculate and plot the ALE main effects of x1, x2, x3, and x4
ALE.1 = ALEPlot(DAT[,2:5], nnet.DAT, pred.fun = yhat, J = 1, K = 500,
NA.plot = TRUE)
ALE.2 = ALEPlot(DAT[,2:5], nnet.DAT, pred.fun = yhat, J = 2, K = 500,
NA.plot = TRUE)
ALE.3 = ALEPlot(DAT[,2:5], nnet.DAT, pred.fun = yhat, J = 3, K = 500,
NA.plot = TRUE)
ALE.4 = ALEPlot(DAT[,2:5], nnet.DAT, pred.fun = yhat, J = 4, K = 500,
NA.plot = TRUE)
```

In the `ALEPlot` implementation, calling the function automatically prints a plot. While this provides some convenience if that is what the user wants, it is not so convenient if the user does not want to print a plot at the very point of ALE creation. It is particularly inconvenient for script building. Although it is possible to configure R to suspend graphic output before the `ALEPlot` is called and then restart it after the function call, this is not so straightforward---the function itself does not give any option to control this behaviour.

ALE interactions can also be calculated and plotted:

```{r ALEPlot nnet ixn}
## Calculate and plot the ALE second-order effects of {x1, x2} and {x1, x4}
ALE.12 = ALEPlot(DAT[,2:5], nnet.DAT, pred.fun = yhat, J = c(1,2), K = 100,
NA.plot = TRUE)
ALE.14 = ALEPlot(DAT[,2:5], nnet.DAT, pred.fun = yhat, J = c(1,4), K = 100,
NA.plot = TRUE)
```

If the output of the `ALEPlot` has been saved to variables, then its contents can be plotted with finer user control using the generic R `plot` method:

```{r ALEPlot nnet organized plots, fig.width=7, fig.height=10}
## Manually plot the ALE main effects on the same scale for easier comparison
## of the relative importance of the four predictor variables
par(mfrow = c(3,2))
plot(ALE.1$x.values, ALE.1$f.values, type="l", xlab="x1",
ylab="ALE_main_x1", xlim = c(0,1), ylim = c(-2,2), main = "(a)")
plot(ALE.2$x.values, ALE.2$f.values, type="l", xlab="x2",
ylab="ALE_main_x2", xlim = c(0,1), ylim = c(-2,2), main = "(b)")
plot(ALE.3$x.values, ALE.3$f.values, type="l", xlab="x3",
ylab="ALE_main_x3", xlim = c(0,1), ylim = c(-2,2), main = "(c)")
plot(ALE.4$x.values, ALE.4$f.values, type="l", xlab="x4",
ylab="ALE_main_x4", xlim = c(0,1), ylim = c(-2,2), main = "(d)")
## Manually plot the ALE second-order effects of {x1, x2} and {x1, x4}
image(ALE.12$x.values[[1]], ALE.12$x.values[[2]], ALE.12$f.values, xlab = "x1",
ylab = "x2", main = "(e)")
contour(ALE.12$x.values[[1]], ALE.12$x.values[[2]], ALE.12$f.values, add=TRUE,
drawlabels=TRUE)
image(ALE.14$x.values[[1]], ALE.14$x.values[[2]], ALE.14$f.values, xlab = "x1",
ylab = "x4", main = "(f)")
contour(ALE.14$x.values[[1]], ALE.14$x.values[[2]], ALE.14$f.values, add=TRUE,
drawlabels=TRUE)

```

### `ale` package equivalent

Now we demonstrate the same functionality with the `ale` package. We will work with the same model on the same data, so we will not create them again.

To create the model, we invoke the `ale` which returns a list with various ALE elements.

```{r ale nnet one-way creation}
library(ale)

nn_ale <- ale(DAT, nnet.DAT, pred_type = "raw")
```

Here are some notable differences compared to `ALEPlot`:

-   In tidyverse style, the first element is the data and the second is the model.
-   Unlike `ALEPlot` that functions on only one variable at a time, `ale` generates ALE data for multiple variables in a dataset at once. By default, it generates ALE elements for all the predictor variables in the dataset that it is given; the user can specify a single variable or any subset of variables. We will cover more details in another vignette, but for our purposes here, we note the `data` element that returns a list of the ALE data for each variable and the `plots` element returns a list of `ggplot` plots.
-   `ale` creates a default generic predict function that matches most standard R models. When the prediction type is not the default "response", as in our case, the user can set the desired type with the `pred_type` argument. However, for more complex or non-standard prediction functions, `ale` supports custom functions with the `pred_fun` argument.

Since the plots are saved as a list, they can easily be printed out all at once:

```{r ale nnet one-way plots, fig.width=3.5, fig.height=10}
# , fig.asp=3 is OK
# Print plots
gridExtra::grid.arrange(grobs = nn_ale$plots, ncol = 1)
```

The `ale` package plots have various features that enhance interpretability:

-   The outcome y is displayed on its full original scale.
-   A median band that shows the middle 5 percentile of the y values is displayed. The idea is that any ALE values outside this band are at least somewhat significant.
-   Similarly, there are 25% and 75% percentile markers to show the middle 50% of the y values. Any ALE y value beyond these bands indicates that the x variable is so strong that it alone at the values indicated can shift the y value by that much.
-   Rug plots indicate the distribution of the data so that outliers are not over-interpreted.

It might not be clear that the previous plots display exactly the same data as those shown above from `ALEPlot`. To make the comparison clearer, we can recalculate the ALEs on a zero-centred scale:

```{r ale nnet one-way zeroed, fig.width=7, fig.height=5}
# Zero-centred ALE
nn_ale <- ale(DAT, nnet.DAT, pred_type = "raw", relative_y = 'zero')

gridExtra::grid.arrange(grobs = nn_ale$plots, ncol = 2)
```

With these zero-centred plots, the full range of y values and the rug plots give some context that aids interpretation. (If the rugs look slightly different, it is because they are randomly jittered to avoid overplotting.)

The `ale` also produces interaction plots. Unlike `ALEPlot`, it does so with a separate dedicated function, `ale_ixn`. By default, it calculates all possible two-way interactions between all variables in the data.

Because the variables interact with each other, the output data structure is a two-layer list, so the print code is slightly more complicated. However, the sample code we provide here using functions from the `purrr` package for iterating through lists and `gridExtra` package for arranging plots are reusable in any application.

```{r ale nnet ixn, fig.width=7, fig.height=10}
# Create and plot interactions
nn_ale_ixn <- ale_ixn(DAT, nnet.DAT, pred_type = "raw")

# Print plots
nn_ale_ixn$plots |>
  purrr::walk(\(.x1) {  # extract list of x1 ALE outputs
    gridExtra::grid.arrange(grobs = .x1, ncol = 1)  # plot all x1 plots
  })
```

These interaction plots are heat maps that indicate the interaction regions that are above or below the average value of y with colours. Grey indicates no meaningful interaction; blue indicates a positive interaction effect; red indicates a negative effect. We find these easier to interpret than the contour maps from `ALEPlot`, especially since the colours in each plot are on the same scale and so the plots are directly comparable with each other.

The range of outcome (y) values is divided into quantiles, deciles by default. However, the middle quantiles are modified. Rather than showing the middle 10% or 20% of values, it is much narrow: it shows the middle 5%. (This value is based on the notion of alpha of 0.05 for confidence intervals; it can be customized with the `median_band` argument.)

The legend shows the midpoint y value of each quantile, which is usually the mean of the boundaries of the quantile. The exception is the special middle quantile, whose displayed midpoint value is the median of the entire dataset.

The interpretation of these interaction plots is that in any given region, the interaction between x1 and x2 increases (blue) or decreases (red) y by the amount indicated over and above the separate individual direct effects of x1 and x2 shown in the one-way ALE plots above. It is **not** an indication of the total effect of both variables together but rather of the additional effect of their interaction- beyond their individual effects. Thus, only the x1-x2 interaction shows any effect. For the interactions with x3, even though x3 indeed has a strong effect on y as we see in the one-way ALE plot above, it has no additional effect in interaction with the other variables, and so its interaction plots are entirely grey.

## Real data with binary outcomes (ALEPlot Example 3)

The next code example from the `ALEPlot` package analyzes a real dataset with a binary outcome variable. Whereas the `ALEPlot` has the user load a CSV file that might not be readily available, we make that dataset available as the census dataset. We load it here with the adjustments necessary to run the `ALEPlot` example.

```{r gbm data setup}
## R code for Example 3
## Load relevant packages
library(ALEPlot)
library(gbm)

## Read data and fit a boosted tree supervised learning model
data(census, package = 'ale')  # load ale package version of the data
data <-  
  census |> 
  as.data.frame() |>   # ALEPlot is not compatible with the tibble format
  select(age:native_country, higher_income) |>  # Rearrange columns to match ALEPlot order
  na.omit(data)
```

Although gradient boosted trees generally perform quite well, they are rather slow. Rather than having you wait for it to run, the code here downloads a pretrained GBM model. However, the code used to generate it is provided in comments so that you can see it and run it yourself if you want to. Note that the model calls is based on `data[,-c(3,4)]`, which drops the third and fourth variables (`fnlwgt` and `education`, respectively).

```{r gbm model}
# To generate the code, uncomment the following lines.
# But it is slow, so this vignette loads a pre-created model object.
# set.seed(0)
# gbm.data <- gbm(higher_income ~ ., data= data[,-c(3,4)],
#                 distribution = "bernoulli", n.trees=6000, shrinkage=0.02,
#                 interaction.depth=3)
# saveRDS(gbm.data, file.choose())
gbm.data <- url('https://github.com/Tripartio/ale/raw/main/download/gbm.data_model.rds') |> 
  readRDS()

gbm.data
```

### ALEPlot code

As before, we create a custom prediction function and then call the `ALEPlot` function to generate the plots. The prediction type here is "link", which represents the log odds in the `gbm` package.

Creation of the ALE plots here is rather slow because the `gbm` predict function is slow. In this example, only `age`, `education_num` (number of years of education), and `hours_per_week` are plotted, along with the interaction between `age` and `hours_per_week`.

```{r ALEPlot gbm, fig.width=7, fig.height=5}
## Define the predictive function; note the additional arguments for the
## predict function in gbm
yhat <- function(X.model, newdata) as.numeric(predict(X.model, newdata,
n.trees = 6000, type="link"))

## Calculate and plot the ALE main and interaction effects for x_1, x_3,
## x_11, and {x_1, x_11}
par(mfrow = c(2,2), mar = c(4,4,2,2)+ 0.1)
ALE.1=ALEPlot(data[,-c(3,4,15)], gbm.data, pred.fun=yhat, J=1, K=500,
NA.plot = TRUE)
ALE.3=ALEPlot(data[,-c(3,4,15)], gbm.data, pred.fun=yhat, J=3, K=500,
NA.plot = TRUE)
ALE.11=ALEPlot(data[,-c(3,4,15)], gbm.data, pred.fun=yhat, J=11, K=500,
NA.plot = TRUE)
ALE.1and11=ALEPlot(data[,-c(3,4,15)], gbm.data, pred.fun=yhat, J=c(1,11),
K=50, NA.plot = FALSE)
```

### `ale` package equivalent

Here is the analogous code using the `ale` package. In this case, we also need to define a custom predict function because of the particular `n.trees = 6000` argument. To speed things up, we provide a pretrained `ale` object. This is possible because `ale` returns objects with data and plots bundled together with no side effects (like automatic printing of created plots). (It is probably possible to similarly cache `ALEPlot` ALE objects, but it is not quite as straightforward.)

#### Log odds

We display all the plots because it is easy to do so with the `ale` package but we focus on `age`, `education_num`, and `hours_per_week` for comparison with ALEPlot. If the shapes of these plots look different, it is because `ale` tries as much as possible to display plots on the same y-axis coordinate scale for easy comparison across plots.

```{r ale one-way link, fig.width=7, fig.height=20}
# Custom predict function that returns log odds
yhat <- function(object, newdata) {
  as.numeric(
    predict(object, newdata,  n.trees = 6000,
            type="link")  # return log odds
  )
}

# Generate ALE data for all variables

# # To generate the code, uncomment the following lines.
# # But it is slow, so this vignette loads a pre-created model object.
# gbm_ale_link <- ale(
#   data[,-c(3,4)], gbm.data,
#   pred_fun = yhat,
#   x_intervals = 500,
#   rug_sample_size = 600,  # technical issue: rug_sample_size must be > x_intervals + 1
#   relative_y = 'zero'  # compatibility with ALEPlot
# )
# saveRDS(gbm_ale_link, file.choose())
gbm_ale_link <- url('https://github.com/Tripartio/ale/raw/main/download/gbm_ale_link.rds') |> 
  readRDS()

# Print plots
gridExtra::grid.arrange(grobs = gbm_ale_link$plots, ncol = 2)
```

Now we generate ALE data for all two-way interactions and then plot them. Again, note the interaction between `age` and `hours_per_week`. The interaction is minimal except for the extremely high cases of hours per week.

```{r ale ixn link, fig.width=7, fig.height=5}
# # To generate the code, uncomment the following lines.
# # But it is slow, so this vignette loads a pre-created model object.
# gbm_ale_ixn_link <- ale_ixn(
#   data[,-c(3,4)], gbm.data,
#   pred_fun = yhat,
#   x_intervals = 500,
#   rug_sample_size = 600,  # technical issue: rug_sample_size must be > x_intervals + 1
#   relative_y = 'zero'  # compatibility with ALEPlot
# )
# saveRDS(gbm_ale_ixn_link, file.choose())
gbm_ale_ixn_link <- url('https://github.com/Tripartio/ale/raw/main/download/gbm_ale_ixn_link.rds') |> 
  readRDS()

# Print plots
gbm_ale_ixn_link$plots |>
  purrr::walk(\(.x1) {  # extract list of x1 ALE outputs
    gridExtra::grid.arrange(grobs = .x1, ncol = 2)  # plot all x1 interaction plots
  })
```

#### Predicted probabilities

Log odds are not necessarily the most interpretable way to express probabilities (though we will show shortly that they are sometimes uniquely valuable). So, we repeat the ALE creation using the "response" prediction type for probabilities and the default median centring of the plots.

As we can see, the shapes of the plots are similar, but the y axes are more easily interpretable as the probability (from 0 to 1) that a census respondent is in the higher income category. The median of around 10% or so indicates the median prediction of the GBM model: half of the respondents were predicted to have higher than a 10% likelihood of being higher income and half were predicted to have lower likelihood. The y-axis rug plots indicate that the predictions were generally rather extreme, either relatively close to 0 or 1, with few predictions in the middle.

```{r ale one-way prob, fig.width=7, fig.height=20}
# Custom predict function that returns predicted probabilities
yhat <- function(object, newdata) {
  as.numeric(
    predict(object, newdata,  n.trees = 6000,
            type="response")  # return predicted probabilities
  )
}

# Generate ALE data for all variables

# # To generate the code, uncomment the following lines.
# # But it is slow, so this vignette loads a pre-created model object.
# gbm_ale_prob <- ale(
#   data[,-c(3,4)], gbm.data,
#   pred_fun = yhat,
#   x_intervals = 500,
#   rug_sample_size = 600  # technical issue: rug_sample_size must be > x_intervals + 1
# )
# saveRDS(gbm_ale_prob, file.choose())
gbm_ale_prob <- url('https://github.com/Tripartio/ale/raw/main/download/gbm_ale_prob.rds') |> 
  readRDS()

# Print plots
gridExtra::grid.arrange(grobs = gbm_ale_prob$plots, ncol = 2)
```

Finally, we again generate two-way interactions, this time based on probabilities instead of on log odds. However, probabilities might not be the best choice for indicating interactions because, as we see from the rugs in the one-way ALE plots, the GBM model heavily concentrates its probabilities in the extremes near 0 and 1. Thus, the plots' suggestions of strong interactions are likely exaggerated. In this case, the log odds ALEs shown above are probably more relevant.

```{r ale ixn prob, fig.width=7, fig.height=5}
# # To generate the code, uncomment the following lines.
# # But it is slow, so this vignette loads a pre-created model object.
# gbm_ale_ixn_prob <- ale_ixn(
#   data[,-c(3,4)], gbm.data,
#   pred_fun = yhat,
#   x_intervals = 500,
#   rug_sample_size = 600  # technical issue: rug_sample_size must be > x_intervals + 1
# )
# saveRDS(gbm_ale_ixn_prob, file.choose())
gbm_ale_ixn_prob <- url('https://github.com/Tripartio/ale/raw/main/download/gbm_ale_ixn_prob.rds') |> 
  readRDS()

# Print plots
gbm_ale_ixn_prob$plots |>
  purrr::walk(\(.x1) {  # extract list of x1 ALE outputs
    gridExtra::grid.arrange(grobs = .x1, ncol = 2)  # plot all x1 plots
  })
```
